\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-02 metatheory-condensed.tex
l/proj2
tighten fmltt figures a bit

W-types are eliminated with the form $\Wrec{\wsig}{\lkg}{t}$, where $t$ is of a
W-type $\El{\wcode{\wsig}}$, and $\lkg$ is essentially an $n$-tuple of
case handlers for the $n$ constructors in~$\wsig$ (\ruleref{tm/wrec}).
Each case handler has a type of the form $\CaseSig{A}{B}{T}$, where
$T$ is the motive of the recursion (\ruleref{ty/casety}).
This collection of case handlers then encodes those defined or inherited
by an \lsti{FRecursion} command in our plugin.
We choose to type it with a linkage type $\TyLkg{\RecSig{\wsig}{T}}$
to avoid introducing $n$-tuples, which linkages generalize.

@@

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the fields preceding the current field~$t$
(i.e., those recorded by the first premise $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$).
Crucially, unlike the second field in a dependent pair (\ruleref{tm/pair}),
the typing of~$t$ in \ruleref{l/add} is polymorphic to how the fields in
$t$'s context are defined, so that a different linkage that overrides
fields in $t$'s context can reuse~$t$---without retyping it---by
projecting it (\ruleref{l/proj2}) and then including it.

@@

We illustrate this with \cref{fig:stlc-linkage-typing}.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{A_i}{t_i}{T_i}$.
%
For readability, we feel free to use a notation with explicit names
in this typing judgment:
we name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
and feel free to write its type, which is a dependent tuple type, as a
dependent record type with labels.
%
\begin{itemize}
  [leftmargin=4ex]
  %[labelsep=*,leftmargin=1pc]

@@

The third premise
$\goodTerm{\Gm,\TyTkg{\lsig}}{\seal}{\sub{A}{\SubstWeak{1}}}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures in~$\lsig$.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
packaging the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new term of type $A$
that hides W-type signatures behind $\cU$, if necessary.
When no hiding is needed, $\seal$ is $\var{0}$ (the identity) and $A$ is
$\TyTkg{\lsig}$.

When a family is closed and a complete linkage~$\lkg$ is formed, fields
of the family can be accessed by projecting them out of the tuple
$\Tkg{\lkg}$, as the last line of \cref{fig:stlc-linkage-typing} suggests.

@@

Inheritance and code reuse can already be expressed through projection
of fields out of linkages and their inclusion into new linkages.
To make common patterns of linkage manipulations more convenient,
\TT builds \emph{linkage transformers}, whose well-formedness judgments
have form $\goodInh{\Gm}{h}{\lsig_1}{\lsig_2}$.
Linkage transformers are second-class; the only way to use them is to
apply them to linkages to form new linkages (\ruleref{tm/inh}).
%
Derived families can be modeled as linkage transformers.
%
Linkage transformers are inductively constructed from several introduction
rules.
\cref{fig:typing-selected} shows the beta rule of an exemplary transformer,
\ruleref{lxeq/ov/beta}.
It can be read as follows: given a linkage of form $\LkgAdd{\lkg}{t_1}$,
applying the transformer $\InhOv{h}{t_2}$ overrides the linkage's last
field~$t_1$ with~$t_2$.
$\InhOv{\InhId}{\wcode{\wsig'_{\texttt{tm}}}}$, for instance, can be used
in \cref{fig:stlc-linkage-typing} to form a derived family that
overrides $\wsig_{\texttt{tm}}$ with an extended
signature~$\wsig'_{\texttt{tm}}$.

@@

The definitive version is formalized in a logical framework with
quotient inductive-inductive types~\cite{nordvall2010inductive,altkap2016},
available as a supplemental file in Agda syntax.
%
We also typeset a more complete set of rules than
\cref{fig:mltt-typing-selected,fig:fmltt-typing-selected} allow
and present it in a supplemental appendix.
%
Another supplemental appendix gives a larger-scale example than
\cref{fig:stlc-linkage-typing} showing how families are modeled and
type-checked using linkages and linkage transformers.

>>>>

W-types are eliminated with the form $\Wrec{\wsig}{\lkg}{t}$, where $t$ is of a
W-type $\El{\wcode{\wsig}}$, and $\lkg$ is essentially an $n$-tuple of
case handlers for the $n$ constructors in~$\wsig$ (\ruleref{tm/wrec}).
Each case handler has a type of the form $\CaseSig{A}{B}{T}$, where
$T$ is the motive of the recursion (\ruleref{tyeq/casety}).
This collection of case handlers then encodes those defined or inherited
by an \lsti{FRecursion} command in our plugin.
We choose to type it with a linkage type $\TyLkg{\RecSig{\wsig}{T}}$
to avoid introducing $n$-tuples, which linkages generalize.

@@

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the fields preceding the current field~$t$
(i.e., those recorded by the first premise $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$).
Crucially, unlike the second field in a dependent pair (\ruleref{tm/pair}),
the typing of~$t$ in \ruleref{l/add} is polymorphic to how the fields in
$t$'s context are defined, so that a different linkage that overrides
fields in $t$'s context can reuse~$t$---without retyping it---by
projecting it (\ruleref{l/proj/2}) and then including it.

@@

\definecolor{A-color}{HTML}{E7EFC5}
\definecolor{T-color}{HTML}{FFF1D0}
\definecolor{t-color}{HTML}{FAE0E4}

We illustrate this with \cref{fig:stlc-linkage-typing}.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{%
\tikzmarkin[disable rounded corners=true,set fill color=A-color,set border color=A-color]{A-i}(-0.02,-0.10)(-0.02,0.28)
A_i%
\tikzmarkend{A-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=t-color,set border color=t-color]{t-i}(-0.02,-0.10)(-0.02,0.28)%
t_i%
\tikzmarkend{t-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=T-color,set border color=T-color]{T-i}(-0.02,-0.10)(-0.02,0.28)%
T_i%
\tikzmarkend{T-i}%
}$.
%
For readability, we feel free to use a notation with explicit names
in this typing judgment:
we name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
and feel free to write its type~$A_i$, which is a dependent tuple type, as a
dependent record type with labels.
%
\begin{itemize}
  [leftmargin=4ex]
  %[labelsep=*,leftmargin=1pc]

@@

The third premise
$\goodTerm{\Gm,\TyTkg{\lsig}}{\seal}{\sub{A}{\SubstWeak{1}}}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
that packages the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is $\var{0}$ (the identity) and $A$ is
$\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are identities,
and $\seal_6$ turns
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$
\endgroup
into
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$%
\endgroup
.

When a family is completed (e.g., \lsti!End!~\lsti!STLC!),
a linkage~$\lkg$ containing all the fields is available
(e.g., $\lkg_{20}$ in \cref{fig:stlc-linkage-typing}).
Fields of the family can then be accessed by projecting them out of the tuple
$\Tkg{\lkg}$.

@@

Inheritance and code reuse can already be expressed through projection
of fields out of linkages and their inclusion into new linkages.
To make common patterns of linkage manipulations more convenient,
\TT provides a ``library'' of \emph{linkage transformers}, whose well-formedness judgments
have form $\goodInh{\Gm}{h}{\lsig_1}{\lsig_2}$. The idea is that
applying~$h$ to a linkage of type~$\TyLkg{\lsig_1}$ yields a linkage of
type~$\TyLkg{\lsig_2}$ (\ruleref{tm/inh}).
%Linkage transformers are second-class; the only way to use them is to
%apply them to linkages to form new linkages.
%
Derived families can be modeled as linkage transformers inductively
constructed from the introduction forms $\InhId$, $\InhExt{h}{t}$, $\InhOv{h}{t}$, etc.
\cref{fig:fmltt-selected} shows the beta rule of an exemplary transformer,
\ruleref{tmeq/ov/beta}.
It can be read as follows: given a linkage of form $\LkgAdd{\lkg}{t_1}$,
applying the transformer $\InhOv{h}{t_2}$ overrides the linkage's last
field~$t_1$ with~$t_2$.
$\InhOv{\InhId}{\wcode{\wsig'_{\texttt{tm}}}}$, for instance, can be used
in \cref{fig:stlc-linkage-typing} to form a derived family that
overrides $\wsig_{\texttt{tm}}$ with an extended
signature~$\wsig'_{\texttt{tm}}$.

@@

The definitive version is formalized in a logical framework with
quotient inductive-inductive types~\cite{nordvall2010inductive,altkap2016},
available as a supplemental file in Agda syntax.
%
We also typeset a more complete set of rules than
the space allows and present it in a supplemental appendix.
%
Another supplemental appendix gives a larger-scale example than
\cref{fig:stlc-linkage-typing} showing how families are modeled and
type-checked using linkages and linkage transformers.
